Water, High-Altitude Condensates, and Methane Depletion in the Atmosphere of the Warm Neptune WASP-107\lowercase{b}

### 2017-08-14 ms.tex
conclusions

The best fit model provides a good fit to the data ($\chi^2_\nu = 1.2$).  There is one $3.4\sigma$ outlier data point (1.34 $\mu$m), so we explored a few retrieval scenarios with additional complexity.  We tested the addition of cloud patchiness (as described by \citealt{line16}) and a quench pressure for nitrogen and carbon species \citep[to account for transport-induced disequilibrium chemistry, e.g.][]{morley17}.  These models did not significantly improve the fit quality, and the main retrieval results were unchanged. We also tested the effect of varying the planet's surface gravity within the published FIXME $\sigma$ range, and found that our results were again minimally changed.

@@

\subsubsection{Methane Abundance}
We ran two additional retrievals to explore whether methane depletion is the underlying cause of the inferred low C/O.  Even though water is the dominant absorber in the WFC3 bandpass, there is enough methane opacity to change the shape of the spectral features at a level we can detect.  In one test, we assumed chemical equilibrium but excluded methane opacity. This set-up resulted in a much broader distribution of C/O values ($0.02 - 1.6$ at $1\,\sigma$), resolving the tension with the solar value.

@@

\subsection{Condensate Properties}
In addition to the atmospheric retrieval, we also performed forward modeling of physically motivated, self-consistent clouds and hazes using the methods described in \cite{fortney08, morley15}.  We model clouds that form in cool atmospheres (Na$_2$S, KCl, ZnS, see \citealt{morley12}), for a range of metallicites from $1-50\times$ solar. We vary the cloud sedimentation efficiency from 0.1 to 3 (highly lofted, small particulate clouds). None of the models produce sufficiently low amplitude spectral features to match the observed spectrum.

@@

%We do not consider ``sloped hazes" at this time as they are unlikely to influence the results.
%We also switch off the quench pressure parameters as they are unlikely to strongly influence the equilibrium abundance profiles of the spectrally prominent, nearly constant-with-altitude abundance, species like CH$_4$, H$_2$O, and NH$_3$ over this temperature range \citep[e.g.][]{line11, moses13}.

\section{Discussion and Conclusions} \label{sec:discuss}
We combined our high precision transmission spectrum of WASP-107b with interior structure modeling to assess the planet's atmospheric composition.

Upper limit on metallicity. The planet's radius is so large that we put an upper limit on the metallicity (additional metals would compress it).  metallicity 30x solar - how does this compare to mass/atmospheric metallicity

Methane depletion
low C/O ratio - planetesimal enrichment (cite Nestor)
internal heat, directly imaged planets

Thick, high-altitude condensate layer

Some directly imaged planets in this temperature regime are also unexpectedly depleted in methane when compared to similar temperature brown dwarfs \citep{skemer14}.

The leading hypothesis of this depletion \citep{zahnle14} is due to the CH4-CO quench point occurring in the CO stability field due to the low gravity of self-luminous planets when compared to brown dwarfs.

In order this mechanism to work for neptune massed planet, the internal temperature would have to be near ~500K to get the deep temperatures high enough to thermochemically deplete methane at the quench point. Given the age of the star, it is unlikely that such high internal temperatures are still due to heat of formation. It is possible that tidal heating could increase the internal temperature (Agundez et al. 2014; Morley et al. 2017) but that would require a negligible eccentricity, of which is currently uncertain for WASP107b.

>>>>

The best fit model provides a good fit to the data ($\chi^2_\nu = 1.2$).  There is one $3.4\sigma$ outlier data point (at 1.34 $\mu$m), so we explored a few retrieval scenarios with additional complexity.  We tested the addition of cloud patchiness (as described by \citealt{line16}) and a quench pressure for nitrogen and carbon species \citep[to account for transport-induced disequilibrium chemistry, e.g.][]{morley17}.  These models did not significantly improve the fit quality, and the main retrieval results were unchanged. We also tested the effect of varying the planet's surface gravity within the published FIXME $\sigma$ range, and found that our results were again minimally changed.

@@

\subsubsection{Methane Abundance}
We ran two additional retrievals to explore whether methane depletion is the underlying cause of the inferred low C/O.  Even though water is the dominant absorber in the WFC3 bandpass, there is enough methane opacity to change the shape of the spectral features at a detectable level for high precision data.  In one test, we assumed chemical equilibrium but excluded methane opacity. This set-up resulted in a much broader distribution of C/O values ($0.02 - 1.6$ at $1\,\sigma$), resolving the tension with the solar value.

@@

\subsection{Condensate Properties}
In addition to the atmospheric retrieval, we also performed forward modeling of physically motivated, self-consistent clouds and hazes using the methods described in \cite{fortney08, morley15}.  We model clouds that form in cool atmospheres (Na$_2$S, KCl, ZnS; see \citealt{morley12}), for a range of metallicites from $1-50\times$ solar. We vary the cloud sedimentation efficiency from 0.1 to 3 (highly lofted, small particulate clouds). None of the models produce sufficiently low amplitude spectral features to match the observed spectrum. FIXME: what did we assume about surface gravity?

@@

%We do not consider ``sloped hazes" at this time as they are unlikely to influence the results.
%We also switch off the quench pressure parameters as they are unlikely to strongly influence the equilibrium abundance profiles of the spectrally prominent, nearly constant-with-altitude abundance, species like CH$_4$, H$_2$O, and NH$_3$ over this temperature range \citep[e.g.][]{line11, moses13}.  
\section{Conclusions} \label{sec:discuss}
We presented an analysis of the atmospheric composition of WASP-107b based on its measured near-infrared transmission spectrum and interior structure modeling of the planet's mass and radius.  Key results from this analysis include:

\begin{itemize}
\item{The upper limit on atmospheric metallicity is $30\times$ solar (assuming the star is young) or FIXME solar for an older system.  Interior structure modeling shows that higher metallicity envelopes would decrease the radius below the observed value.  Based on the Solar System trend of increasing planet metallicity with decreasing planet mass, WASP-107b is expected to have a metallicity of $30\times$ solar, just at the edge of our allowed range \citep{kreidberg14b}. Further characterization of the stellar age is needed to determine whether WASP-107b is consistent with expectations, or discrepant with the Solar System trend, which would suggest an intriguing difference in its formation history.}
\item{The methane abundance is depleted relative to expectations for a solar abundance pattern. This could be due to an instrinsically low carbon-to-oxygen ratio, which could arise from accretion of water-rich planetesimals \citep{mordasini16, espinoza17}.   Another possibility is that the planet has a hot interior.}
\item{Optically thick condensates are present at high altitudes ($0.1 - 3$ mbar).  Existing models of physically-motivated cloud and haze formation do not satisfactorily reproduce the observed . Clouds form too deep in the atmosphere }
\end{itemize}

%Some directly imaged planets in this temperature regime are also unexpectedly depleted in methane when compared to similar temperature brown dwarfs \citep{skemer14}.   The leading hypothesis of this depletion \citep{zahnle14} is due to the CH4-CO quench point occurring in the CO stability field due to the low gravity of self-luminous planets when compared to brown dwarfs.

%In order this mechanism to work for neptune massed planet, the internal temperature would have to be near ~500K to get the deep temperatures high enough to thermochemically deplete methane at the quench point. Given the age of the star, it is unlikely that such high internal temperatures are still due to heat of formation. It is possible that tidal heating could increase the internal temperature (Agundez et al. 2014; Morley et al. 2017) but that would require a negligible eccentricity, of which is currently uncertain for WASP107b.
